---
title: "Microarray Pipeline: Read data, quality control, histrogram/microarray chip images, save raw data"
author: Sherlyn Jemimah, PhD
output: html_notebook
---

# Pre-requisites: libraries and data
You need the following Bioconductor packages for Affymetrix array analysis:
affy, affyPLM and simpleaffy if you want with older arrays (3' arrays)
oligo if you work with newer arrays (HTA, Gene ST...)

It is best to pick one of these two choices. If you load multiple packages with similar functionality, e.g. both affy and oligo or both simpleaffy and oligo, R will become confused. The affy-based packages and oligo contain methods with the same name (e.g. intensity(), MAplot(), rma()) but with slightly different code. So R will not know if it has to use the intensity() method from the affy packages or from oligo. By default it chooses affy over oligo. So you always have to specify the packagename for the oligo methods (see section on how to specify the package of a method) and even then it does not always work well.

affydata if you need example data sets
ArrayExpress if you want to obtain data from ArrayExpress
limma
Biobase
Biostrings
genefilter
Bioconductor packages for the annotation of the spots on the arrays

## Load libraries
```{r}
library(affy)
library(Biobase)
library(Biostrings)
library(dplyr)
library(GEOquery)
library(genefilter)
library(limma)
library(tibble)
#library(org.Hs.eg.db)
#library(hgu133plus2.db)
library(hgu133plus2cdf)
#library(oligo)
```

## Load data
```{r}
my.gse <- "GSE10846"
setwd("D:/microarray/")
if(!file.exists("geo_downloads")) dir.create("geo_downloads")
if(!file.exists("results"))  dir.create("results", recursive=TRUE)
my.geo.gse <- getGEO(GEO=my.gse, filename=NULL, destdir="geo_downloads", GSElimits=NULL, GSEMatrix=TRUE, AnnotGPL=TRUE, getGPL=TRUE)
##get rid of list structure
my.geo.gse <- my.geo.gse[[1]]
```

## Untar CEL files
```{r}
if(!file.exists(paste0("./geo_downloads/",my.gse)))
getGEOSuppFiles(my.gse, makeDirectory=T, baseDir="geo_downloads")
untar(paste0("geo_downloads/",my.gse,"/",my.gse,"_RAW.tar"), exdir=paste0("geo_downloads/",my.gse,"/CEL"))
# get file list
my.cels <- list.files(paste0("geo_downloads/",my.gse,"/CEL"), pattern=".cel", ignore.case = T)
my.cels <- sort(my.cels)
```

# Format Phenodata: modify according to array specifics
```{r}
##make data frame of phenoData
my.pdata <- as.data.frame(pData(my.geo.gse), stringsAsFactors=F)
## reformat
# gender
my.pdata$characteristics_ch1 <- substring(my.pdata$characteristics_ch1, 9) # remove "Gender: " from string
my.pdata <- dplyr::rename(my.pdata, Gender=characteristics_ch1)
# age
my.pdata$characteristics_ch1.1 <- substring(my.pdata$characteristics_ch1.1, 6) # remove "Age: " from string
my.pdata <-  dplyr::rename(my.pdata, Age=characteristics_ch1.1)
# DLBC_class
my.pdata$characteristics_ch1.6 <- substring(my.pdata$characteristics_ch1.6, 44) # remove "Clinical info: Final microarray diagnosis: "
my.pdata <- rename(my.pdata, DLBC_Class=characteristics_ch1.6)
my.pdata$DLBC_Class <- gsub(' DLBCL', '', my.pdata$DLBC_Class)
# status
my.pdata$characteristics_ch1.7 <- substring(my.pdata$characteristics_ch1.7, 34) # remove "Clinical info: Follow up status: "
my.pdata <-  dplyr::rename(my.pdata, Status=characteristics_ch1.7)
# status
my.pdata$characteristics_ch1.8 <- substring(my.pdata$characteristics_ch1.8, 33) # remove "Clinical info: Follow up years: "
my.pdata <-  dplyr::rename(my.pdata, Follow_up_years=characteristics_ch1.8)
my.pdata$Follow_up_years <- as.numeric(my.pdata$Follow_up_years)
# Treatment
my.pdata$characteristics_ch1.9 <- substring(my.pdata$characteristics_ch1.9, 30) # remove "Clinical info: Chemotherapy: "
my.pdata <-  dplyr::rename(my.pdata, Treatment=characteristics_ch1.9)
my.pdata$Treatment <- gsub('-Like Regimen', '', my.pdata$Treatment)
# ECOG
my.pdata$characteristics_ch1.10 <- substring(my.pdata$characteristics_ch1.10, 41) # remove "Clinical info: ECOG performance status: "
my.pdata <-  dplyr::rename(my.pdata, ECOG=characteristics_ch1.10)
# Stage
my.pdata$characteristics_ch1.11 <- substring(my.pdata$characteristics_ch1.11, 23) # remove "Clinical info: Stage: " from string
my.pdata <-  dplyr::rename(my.pdata, Stage=characteristics_ch1.11)

## remove unnecessary columns
my.pdata <- my.pdata[, c("Gender", "Age", "DLBC_Class", "Follow_up_years", "Status", "Treatment", "ECOG", "Stage")]
my.pdata <- my.pdata[order(rownames(my.pdata)), ]

## manual tweaking of rownames
rownames(my.pdata) <- paste(rownames(my.pdata), ".cel.gz", sep="")
rownames(my.pdata) <- gsub("GSM361244.cel.gz","GSM361244.CEL.gz",rownames(my.pdata))
rownames(my.pdata) <- gsub("GSM361243.cel.gz","GSM361243.CEL.gz",rownames(my.pdata))
rownames(my.pdata) <- gsub("GSM361242.cel.gz","GSM361242.CEL.gz",rownames(my.pdata))
rownames(my.pdata) <- gsub("GSM361241.cel.gz","GSM361241.CEL.gz",rownames(my.pdata))
rownames(my.pdata) <- gsub("GSM361240.cel.gz","GSM361240.CEL.gz",rownames(my.pdata))
rownames(my.pdata) <- gsub("GSM361239.cel.gz","GSM361239.CEL.gz",rownames(my.pdata))

## check that the filenames and phenodata are exactly same
rownames(my.pdata) == my.cels
write.table(my.pdata, file=paste0("geo_downloads/",my.gse,"/CEL/",my.gse,"_SelectPhenoData.txt"), sep="\t", quote=F)
```
# Read CEL files using affy
```{r}
cel.path <- paste0("geo_downloads/",my.gse,"/CEL")
data <- ReadAffy(celfile.path=cel.path,phenoData=paste(cel.path, paste0(my.gse,"_SelectPhenoData.txt"), sep="/"))
```

