---
title: "Microarray Pipeline"
output: html_notebook
---

### Inputs required

1. GEO Accession ID
2. Phenotype file (list of samples and descriptors)
```{r}
library(GEOquery)
library(dplyr)
library(limma)
library(RColorBrewer)
library(affy)
library(tibble)

# input GEO ID
my.gse <- "GSE31312"
# set working directory
wd <- "D:/microarray/"
setwd(wd)
```

Load file into memory
```{r}
if(!file.exists("geo_downloads")) dir.create("geo_downloads")
if(!file.exists("results"))  dir.create("results", recursive=TRUE)
my.geo.gse <- getGEO(GEO=my.gse, filename=NULL, destdir="geo_downloads", GSElimits=NULL, GSEMatrix=TRUE, AnnotGPL=TRUE, getGPL=TRUE)
##get rid of list structure
my.geo.gse <- my.geo.gse[[1]]
```

Optional: The following chunk enables you to check the microarray data structure. If you have already performed this step in previous iterations, you can skip this.
```{r}
class(my.geo.gse)
names(my.geo.gse)
length(my.geo.gse)
##object is an ExpressionSet
class(my.geo.gse)
##you can see the structure of the object
str(my.geo.gse)
colnames(pData(my.geo.gse))
pData(my.geo.gse)$data_processing[1]
head(exprs(my.geo.gse))
summary(exprs(my.geo.gse))
```

Get file list
```{r}
if(!file.exists(paste0("./geo_downloads/",my.gse)))
getGEOSuppFiles(my.gse, makeDirectory=T, baseDir="geo_downloads")
list.files(paste0("geo_downloads/",my.gse))
```

Decompress the files.
```{r}
untar(paste0("geo_downloads/",my.gse,"/",my.gse,"_RAW.tar"), exdir=paste0("geo_downloads/",my.gse,"/CEL"))
list.files(paste0("geo_downloads/",my.gse,"/CEL"))
```

```{r}
my.cels <- list.files(paste0("geo_downloads/",my.gse,"/CEL"), pattern=".cel", ignore.case = T)
my.cels <- sort(my.cels)
```

### Process phenodata
```{r}
##make data frame of phenoData
my.pdata <- as.data.frame(pData(my.geo.gse), stringsAsFactors=F)
head(my.pdata)
```

For GSE31312
```{r}
# DLBC_class
my.pdata$characteristics_ch1 <- substring(my.pdata$characteristics_ch1, 37) # remove "gene expression profiling subgroup: "
my.pdata <- rename(my.pdata, DLBC_Class=characteristics_ch1)
# Treatment
my.pdata$treatment_protocol_ch1 <- substring(my.pdata$treatment_protocol_ch1, 128) # 
my.pdata <- rename(my.pdata, Treatment=treatment_protocol_ch1)
my.pdata$Treatment <- gsub(').', '', my.pdata$Treatment)
```

For GSE10846
```{r}
# gender
my.pdata$characteristics_ch1 <- substring(my.pdata$characteristics_ch1, 9) # remove "Gender: " from string
my.pdata <- rename(my.pdata, Gender=characteristics_ch1)
# age
my.pdata$characteristics_ch1.1 <- substring(my.pdata$characteristics_ch1.1, 6) # remove "Age: " from string
my.pdata <- rename(my.pdata, Age=characteristics_ch1.1)
# DLBC_class
my.pdata$characteristics_ch1.6 <- substring(my.pdata$characteristics_ch1.6, 44) # remove "Clinical info: Final microarray diagnosis: "
my.pdata <- rename(my.pdata, DLBC_Class=characteristics_ch1.6)
my.pdata$DLBC_Class <- gsub(' DLBCL', '', my.pdata$DLBC_Class)
# status
my.pdata$characteristics_ch1.7 <- substring(my.pdata$characteristics_ch1.7, 34) # remove "Clinical info: Follow up status: "
my.pdata <- rename(my.pdata, Status=characteristics_ch1.7)
# status
my.pdata$characteristics_ch1.8 <- substring(my.pdata$characteristics_ch1.8, 33) # remove "Clinical info: Follow up years: "
my.pdata <- rename(my.pdata, Follow_up_years=characteristics_ch1.8)
my.pdata$Follow_up_years <- as.numeric(my.pdata$Follow_up_years)
# Treatment
my.pdata$characteristics_ch1.9 <- substring(my.pdata$characteristics_ch1.9, 30) # remove "Clinical info: Chemotherapy: "
my.pdata <- rename(my.pdata, Treatment=characteristics_ch1.9)
my.pdata$Treatment <- gsub('-Like Regimen', '', my.pdata$Treatment)
# ECOG
my.pdata$characteristics_ch1.10 <- substring(my.pdata$characteristics_ch1.10, 41) # remove "Clinical info: ECOG performance status: "
my.pdata <- rename(my.pdata, ECOG=characteristics_ch1.10)
# Stage
my.pdata$characteristics_ch1.11 <- substring(my.pdata$characteristics_ch1.11, 23) # remove "Clinical info: Stage: " from string
my.pdata <- rename(my.pdata, Stage=characteristics_ch1.11)
```

Remove unnecessary columns and match with CEL file names: GSE10846
```{r}
dim(my.pdata)
colnames(my.pdata)
head(my.pdata[, c("Gender", "Age", "DLBC_Class", "Follow_up_years", "Status", "Treatment", "ECOG", "Stage")], 10)
my.pdata <- my.pdata[, c("Gender", "Age", "DLBC_Class", "Follow_up_years", "Status", "Treatment", "ECOG", "Stage")]
my.pdata <- my.pdata[order(rownames(my.pdata)), ]
head(my.pdata, 10)
```

Remove unnecessary columns and match with CEL file names: GSE31312
```{r}
dim(my.pdata)
colnames(my.pdata)
head(my.pdata[, c("DLBC_Class", "Treatment")], 10)
my.pdata <- my.pdata[, c("DLBC_Class", "Treatment")]
my.pdata <- my.pdata[order(rownames(my.pdata)), ]
head(my.pdata, 10)
```

```{r}
##the rownames of the data frame must be the same as the CEL file names.
##For this data, we need to do a bit of tinkering.
head(my.cels)
my.pdata <- paste(rownames(my.pdata), ".CEL.gz", sep="")
my.pdata %>% 
  rownames_to_column(data=my.pdata,var="sample") %>%
  dplyr::filter(sample %in% my.cels) %>%
  column_to_rownames(data=my.pdata,var="sample")
temp.rownames <- rownames(my.pdata)
rownames(my.pdata) == my.cels
table(temp.rownames == my.cels)
rownames(my.pdata) <- temp.rownames
# rownames(my.pdata) <- gsub("GSM361244.cel.gz","GSM361244.CEL.gz",rownames(my.pdata))
# rownames(my.pdata) <- gsub("GSM361243.cel.gz","GSM361243.CEL.gz",rownames(my.pdata))
# rownames(my.pdata) <- gsub("GSM361242.cel.gz","GSM361242.CEL.gz",rownames(my.pdata))
# rownames(my.pdata) <- gsub("GSM361241.cel.gz","GSM361241.CEL.gz",rownames(my.pdata))
# rownames(my.pdata) <- gsub("GSM361240.cel.gz","GSM361240.CEL.gz",rownames(my.pdata))
# rownames(my.pdata) <- gsub("GSM361239.cel.gz","GSM361239.CEL.gz",rownames(my.pdata))
rm(temp.rownames)
table(rownames(my.pdata) == my.cels)

```

Write tables
```{r}
write.table(my.pdata, file=paste0("geo_downloads/",my.gse,"/CEL/",my.gse,"_SelectPhenoData.txt"), sep="\t", quote=F)
```


Get raw counts
```{r}
cel.path <- paste0("geo_downloads/",my.gse,"/CEL")
my.affy <- ReadAffy(celfile.path=cel.path, phenoData=paste(cel.path, paste0(my.gse,"_SelectPhenoData.txt"), sep="/"))
show(my.affy)
head(exprs(my.affy))
dim(exprs(my.affy))
colnames(pData(my.affy))
pData(my.affy)
pData(my.affy)$title
```

```{r}
my.rma <- rma(my.affy, normalize=F, background=F)
plotDensity(exprs(my.rma))
```

```{r}
boxplot(exprs(my.affy), las=2, outline=F, main="Arrays Not Normalized")
```

