---
title: "Cox Proportional Hazards analysis"
author: Sherlyn Jemimah, PhD
output: html_notebook
---

# Load data
Load clinical data and append gene expression data to the dataframe
```{r}
library(dplyr)
library(tibble)
library(reshape)
library(tidyverse)

# genes in HR pathway
HR_genes <- c("BLM", "RMI1", "TOP3A", "TOP3B", "MRE11A", "NBN", "RAD50", "PPP4C", "PPP4R1", "PPP4R2", "PPP4R4", "RAD52", "RDM1", "RAD54L", "RAD51", "RAD51B", "RAD51C", "RAD51D", "RECQL", "RECQL5", "WRN", "RPA1", "RPA2", "RPA3", "RPA4", "EID3", "NDNL2", "NFATC2IP", "NSMCE1", "NSMCE2", "NSMCE4A", "PAXIP1", "SMC5", "SMC6", "BRCA1", "BRCA2", "EME1", "EME2", "GEN1", "KAT5", "MUS81", "PARG", "RAD54B", "SHFM1", "SLX1A", "SLX4", "SPO11", "UIMC1", "TOPBP1", "POLD1", "POLD2", "POLD3", "POLD4", "XRCC3", "XRCC2", "DNA2", "CTIP", "CTCF", "SSBP1")

# gene expression data
expdata <- read.table("D:/Shristi/GSE10846_Genes.csv",sep=",",header=T)
rownames(expdata) <- expdata$ï..Gene
expdata <- subset(expdata, select = -c(ï..Gene) )
# filter selected genes
expdata <- expdata %>%
  dplyr::filter(row.names(expdata) %in% HR_genes)

# clinic data
clindata <- read.table("D:/Shristi/GSE10846_Clinical.csv",sep=",",header=T)
rownames(clindata) <- clindata$ID_REF

# append exp data to clinical data
get_exp <- function(x,gene){ expdata[gene,x] }
for (gene in HR_genes){
  clindata[gene] <- sapply(clindata$ï..ID_REF,get_exp,gene=gene)
}
```

# Cox Proportional Hazards (CPH) Model: univariate
Install required libraries and call the CPH regression model in R
```{r}
## if the packages are not installed:
# install.packages(c("survival","survminer"))

library(survival)
library(survminer)
res.cox <- coxph(Surv(as.numeric(OS_time),as.numeric(OS_status)) ~ RAD51, data=clindata)
summary(res.cox)
```
# Notes
1. “z” gives the Wald statistic value
2. Sign of the regression coefficients (coef): A positive sign means that the hazard (risk of death) is higher, and thus the prognosis worse, for subjects with higher values of that variable

# Cox Proportional Hazards (CPH) Model: multivariate
```{r}
## if the packages are not installed:
# install.packages(c("survival","survminer"))

library(survival)
library(survminer)
res.cox <- coxph(Surv(as.numeric(OS_time),as.numeric(OS_status)) ~ RAD51+BLM+PPP4C, data=clindata)
summary(res.cox)
```

# Visualization
```{r}
# Plot the baseline survival function
ggsurvplot(survfit(res.cox), color = "#2E9FDF",
           ggtheme = theme_minimal(), data=clindata)
```

```{r}
clindata$R51q <- ntile(clindata[,"RAD51"], n=2)
clindata <- clindata[!is.na(clindata$R51q), ]
lowR <- survfit(Surv(as.numeric(OS_time),as.numeric(OS_status)) ~ RAD51, data=clindata[clindata$R51q==1,])
higR <- survfit(Surv(as.numeric(OS_time),as.numeric(OS_status)) ~ RAD51, data=clindata[clindata$R51q==2,])
fit  <- list(Low=lowR, High=higR)
ggsurvplot_combine(fit,clindata)
```


